---
title: |
  <center> Greatness in Sports: </center>
  <center> Understanding and Visualizing its Patterns </center>
author: "Benjamin Leinwand and Matthew Murray"
geometry: "top = 2cm, bottom = 2cm, left = 2cm, right = 2cm"
fontsize: 10 pt
bibliography: references.bib
output: 
  pdf_document:
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, options(knitr.table.format = "markdown"))
```

```{r code, include=FALSE, warning=FALSE}


##


library(janitor)
library(dplyr)
library(tidyverse)
library(lubridate)
library(stringi)
library(r2r)
library(purrr)
library(scales)
library(broom)
library(kableExtra)


###


# read-in
data <- read.csv("data.csv")
data.2 <- read.csv("data.2.csv")

# truncate
data <- data[data$Year >= 1945, ]

# aggregate
data.agg <- data %>% group_by(Year) %>% summarise(N = sum(N))


###


# filter data for years, sports with no GOAT candidates
data.0 <- data[data$N == 0,]

# left merge
data.combined <- merge(x = data.0, y = data.2, by = "Sport", all.x = TRUE)
data.combined.2 <- data.combined %>%
  group_by(Sport) %>%
  summarize(N= n(), avg.length = unique(avg.length))

# add new indicator variable for contact sports
data.combined.2$Contact <- "No"
data.combined.2[c(2, 4, 6, 7), 4] <- "Yes"


###


fig.12 <- ggplot(data = data.combined.2, 
       aes(x = avg.length, y = as.integer(N),
           shape = Sport, color = Contact)) +
  scale_shape_manual(values = seq(10)) + 
  scale_x_continuous(expand = c(0, 0), limits = c(0, 25)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 80)) +
  geom_point() +
  labs(x = "Average Career Length in Years", 
       y = "# Years with no GOAT Candidates", 
       title = "# Years with No Potential GOATs vs. Average Career Length for GOAT Candidates",
       caption = "Figure 12") + 
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  theme(plot.title=element_text(family = "serif", face = "bold", hjust = 0.5, size = 12)) +
  labs(colour = "Contact Sport?") 


```


\textcolor{red}{Note: Limitations/concerns are noted in red.}

\textcolor{red}{We are aware and cognizant of the fact that if a reader were to conduct a similar analysis, the results and data may be very different. However, our goal in this project is not to conduct an analysis that is extremely robust/resistant to sensitivty tests, but rather, one that is informative, insightful, and interesting. If 25 different researchers were to conduct this analysis, each dataset and approach may be very different, and we are okay with that fact.}

# 1 Introduction

## 1.1 Background

A common phrase that is emphasized among sports fans and expert calls to not take greatness for granted. While the sports themselves are (hopefully) lasting, the athletes come and go, with their careers seeming like a brief blip in time when it is all said and done. Baseball fans listen to stories of the Great Bambino from their elders, wondering if they will ever see a player or a talent burst on to the scene again. Meanwhile, basketball fans seem to engage in an everlasting debate between Michael Jordan and Lebron James. 

Nonetheless, with the seemingly endless abundance of talent on display today, coupled with the advances being made in athlete performance and recovery that have athletes performing at the highest levels ever, one has to call to question: can we take greatness for granted? Are there more greatest of all time (GOAT) candidates than one may intuitively believe there to be? Are there patterns in which greatness occurs in one sport, and a lack thereof in another? Do certainly qualities (e.g: contact vs. non-contact) tend to attract more GOAT candidates to a sport than to another. In this paper, we seek to provide the answers to these questions. 


## 1.2 Data Collection

### 1.2.1 Sports 

Due to the unique nature of this project, we had to impute our own datasets. We decided to compile lists of men's GOAT candidates from the following sports:

***"Subjective" Sports***
1. Soccer
2. Cricket
3. Tennis
4. Basketball
5. Baseball
6. Football
7. Ice Hockey
8. Golf

***"Objective" Sports***
9. Chess
10. Swimming
11. Running 

The inclusion of football, baseball, basketball, and ice hockey can be attributed to the fact that they are the 4 of the 5 most popular sports in the United States [@Gallup]. Golf and tennis were included because professionals in both sports compete in four "major" tournaments each year. Soccer and cricket were included due to their immense popularity and following worldwide [@SFB]. 

Additionally, we decided to impute data for three "objective" sports - chess, swimmming, and running. Put simply, these sports have very well-accepted, objective metrics that are more-than-often referred to when discussing the best player in that sport.

### 1.2.2 GOAT Sources and Rationale

For each "subjective" sport, except golf, we imputed two lists that rank the best players of all time. One list would be a subjective ranking of the players from a reputable, respected source. While the reputability of a source is inherently subjective, we tried as much as possible to include sources that are mainstream and whose respectability would be difficult to refute. The sources of the subjective ranking lists for each sport are as follows:

* Soccer - *Sports Illustrated*
* Cricket - *British Broadcasting Company (BBC)*
* Tennis - *Stadium Talk*
* Basketball - *ESPN*
* Baseball - *ESPN*
* Football - *ClutchPoints*
* Ice Hockey - *Bleacher Report*
* Golf - None

The other list would be based on an objective metric, such as the number of MVPs won throughout one's career or a comprehensive statistic that aptly captures a players value. We tried to use metrics that are fair, comprehensive, and holistic in that they thoroughly and aptly capture a player's greatness. The criteria/measures used for all sports - both subjective and objective- as well as the inevitable limitation(s) of each criteria - are noted below:

* Soccer - Players were ranked by the number of Ballon d'Ors they won. More specifically, the list includes the number of Ballon d'Ors won by each player *since 2007*. The Ballon d'Or was chosen because it is widely recognized as the most prestigious award in soccer. Nonetheless, this award does have a few limitations. From 1956 (the first year the award was given out) to 1995, the Ballon d'Or was only given to the best soccer player of European origin. In 1995, the award was expanded to consider soccer players from any European club. It was not until 2007 that the award was expanded to consider all soccer players worldwide [@CBC]. 

* Cricket - Players were ranked based on their batting average. It is equal to the number of runs a player has scored divided by the number of time he has gotten out; in other words, it is the average number of runs a player scores at bat before getting out. We arbitrarily chose batting average because it is one of the most commonly cited offensive statistics for cricket. (For instance, Donald Bradman is often noted as the greatest cricketers of all time *because* of his ridiculously high batting average.) Nonetheless, a limitation of using this metric is that it does not account for great cricket bowlers. 

* Tennis - Players were ranked based on the number of majors they won. The four majors, also known as Grand Slam events, in tennis include the Australian Open, the French Open, Wimbledon, and the US Open, each of which is played annually. These tournaments became open to professionals in 1968, thus marking the advent of the Open Era in tennis as well as a significant shift and maturation of the sport [@TC]. Today, tennis player accomplishments are often discussed in the context of Open Era records, with one of the most prominent ones being Grand Slam or major titles. 

* Basketball - Players were ranked based on their Value over Replacement Player (VORP). This metric measures "the number of points that a player contributed per 100 team possessions above a replacement level player" [@Qlik].

* Baseball - Players were ranked based on their Most Valuable Player (MVP) shares. Put simply, an award share is equal to the number of points that a player received for an award divided by the total number of first place votes. For example, assuming there are 10 votes, if a player wins 9 first place votes, each of which are worth 3 points, and 1 second place vote, which is worth 2 points, the player wins up with an award share of 29/30 ($\approx$ 0.967). The benefit of using this metric is that it rewards and considers players who did not necessarily win the MVP, but were close. 

* Football - Players were ranked based on the number of MVP Awards they won. More specifically, the list includes players who won more than 1 NFL MVP.  

* Ice Hockey - Players were ranked based on their points. A player earns 1 point every time he either scores or assists a goal. We arbitrarily chose chose points because it is one of the most commonly cited offensive statistics in hockey. (For instance, Wayne Gretzky is often noted as the greatest hockey player of all time *because* he has the most points among all players even if one were to take away his goals scored.) Nonetheless, a limitation of using this metric is that it does not account for defencemen and goalies, whose values cannot necessarily be deciphered through points. 

* Golf - Players were ranked based on the number of majors (Masters, PGA Championship, U.S. Open, and The Open) they won. More specifically, we included players who have won at least 5 majors. Additionally, we ranked players based on PGA tournament victories.

Additionally, we decided to impute data for three "objective" sports. Put simply, these sports have very well-accepted, objective metrics that are more-than-often referred to when discussing the best player in that sport. The objective sports and corresponding metrics are as follows:

* Chess - Players were ranked based on their ELO rating. Like any metric, the ELO rating quantifies the relative strength and skill of a chess players based on his or her previous matches. More information about the metric, including its formula, can be found [here](https://chessklub.com/elo-rating-system/). We included players who have held the all-time record for the ELO rating at a point in time. While the ELO rating system is widely used today, the main limitation of this metric is that it was not adopted by the Federation International des Echecs (FIDE) until 1970 [@ChessKlub]. In other words, there were no "official" chess ratings before 1970. Consequently, to include chess greats from 1945 to 1970, we also decided to include and based on their Chessmetrics performance rating. More specifically, as with the ELO rating, we included players who have held the all-time record for the Chessmetrics rating at a point in time. Designed by Jeff Sonas, the Chessmetrics rating system was created with the intention of improving upon the slow and conservative nature of the ELO rating system [@ChessMetrics]. The rating system has also been used to retroactively rate players who played before 1970 (before the adoption of ELO by the FIDE). More information about the rating can be found [here](http://www.chessmetrics.com/cm/CM2/Formulas.asp?Params=).

* Swimming - Swimmers who have held the record for the 100 Meter Freestyle race at a point in time were included in the dataset. While there are many swimming races and events to choose from, this event was chosen because it is often considered to be the most prominent, blue-ribbon event in the sport of swimming. We also decided to include Michael Phelps due to the fact that he has 23 Olympic gold medals, which is more than twice that of the Olympic swimmer with the second most Olympic gold medals (Mark Spitz).

### 1.2.3 Time Span

Each of our graphs includes data/players from 1945 to the present (although we may later change this to 2010; after 2010, we are likely underestimating the number of GOATs just because most of the great players have not been playing long enough to truly be in the GOAT conversation; a perfect example of such a player would be Patrick Mahomes II, who plays for the Kansas City Chiefs). The choice of 1945 can be attributed to the fact that at this time, the MLB (1876) and NFL (1920) were already founded, while the Basketball Association of America (which eventually became the NBA) was founded shortly thereafter in 1946. Therefore, one can argue that the three most popular sports in North America (in addition to soccer) were already up and running by that time period. Additionally, 1945 marked the end of World War II and the advent of a period of economic flourishing in the Western world. 

### 1.2.4 Career Span

Another challenge that we encountered during the data imputation process was properly, consistently, and logically defining defining the beginning and end of a player's professional career. In doing so, we hoped to capture the **entirety** of a player's career, including the advent, peak, and twilight of one's playing years. The specific details of how we defined the beginning and end of a player's career is below:

* Soccer - We imputed a player's active years as the first and last calendar years that one played professionally, either internationally *or* for a professional club), and all years in between. 

* Cricket - We imputed a player's active years in 1 of 2 ways. If the player was from our arbitrary list, the active years equate to the first and last calendar years that one played professionally - either internationally *or* domestically - and all years in between. If the player was from our list based on batting average, the active years equate to the first and last year he played in a test match, and all years in between. In the case that a player is on both lists, we used the former method. 

* Tennis - We imputed a player's active years as the first and last calendar years that one played professionally 

* Basketball - We imputed a player's active years as all years that he played in the NBA. 

* Baseball - We imputed a player's active years as all years that he played in the MLB. 

* Football - We imputed a player's active years as all years that he played in the NFL.

* Ice Hockey - We imputed a player's active years as all years that he played in the NHL.

* Golf - We imputed a player's active years as the first and last calendar years that he won a major, and all years in between. For example, since Tiger Woods won his first major in 1997 and his last major in 2019, his active years are listed as 1997-2019, even though he is still playing in professional golf tournaments (in fact I know he most recently played in the Genesis Open). The reason why we used this logic is that professional golfers can play for a longer (compared to athletes in other sports) period of time due to the presence of the PGA Tour Champions (formerly known as the Senior PGA Tour) as well as the nature of the sport itself (more specifically, the fact that players can sit out certain tournaments and the less violent nature of the sport). For example, Jack Nicklaus, whom many consider to be the greatest golfer of all time, played at his last U.S. Open in 2000, but did not play his last Masters, PGA Championship, and Open Championship until 2005 (add source here). 

* Chess - We imputed a player's active years as the first and last years that he played in a competitive tournament. 

* Swimming - We imputed an athlete's active years as the first and last years that he computed in a major swimming tournament. While their is inherent subjectivity in what is considered a "major" tournament and what is not, we counted major swimming events such as the Olympics, European Aquatics Championships, World Short Course Championships, and FINA Swimming World Cup, to name a few. Put simply, tournaments in which results and data are readily available are presumed to be "major". 

\textcolor{red}{Another limitation to take into account is the fact that we imputed data year-by-year for athletes in some sports, but not in others. A potential result of this limitation is that said years may have an inflated number of active GOAT candidates.}

\newpage

# Topics of Interest 


## Contact vs. Non-Contact Sports

One distinction that we chose to investigate is contact sports vs. non-contact sports. An article published by *The American Academy of Pediatrics* titled "Medical Conditions Affecting Sports Participation" divides sports into three main categories: (1) contact, (2) limited-contact, and (3) non contact. Furthermore, contact sports are sub-divided into (1) collision and (2) contact sports. It further states:

"In collision sports (eg, boxing, ice hockey, football, lacrosse, and rodeo), athletes purposely hit or collide with each other or with inanimate objects (including the ground) with great force. In contact sports (eg, basketball and soccer), athletes routinely make contact with each other or with inanimate objects but usually with less force than in collision sports. In limited-contact sports (eg, softball and squash), contact with other athletes or with inanimate objects is infrequent or inadvertent. However, some limited-contact sports (eg, skateboarding) can be as dangerous as collision or contact sports. Even in noncontact sports (eg, power lifting), in which contact is rare and unexpected, serious injuries can occur." 

Using the criteria noted above, we decided to classify **basketball, football, hockey, and soccer** as **contact** sports; conversely, **baseball, cricket, golf, and tennis** were deemed **limited contact or non contact**. In doing so, we were looking to investigate whether the type of sport that a player participates in affects a GOAT candidate's career length. 

```{r}

fig.12

```


Based on **Figure 11**, no trends in particular seem to stick out. Although football and basketball (both of which are contact sports) both seem to have shorter average career lengths, one cannot say this for all contact sports compared to non-contact sports. However, a potential limitation/inconsistency for golf (a non-contact sport) is that the average career length may appear to be shorter than it actually is because we used the first and last years a player won a major for a career span.

\newpage

## Maximal Value for Increasing Distribution

One can examine the idea of finding the next GOAT candidate from a probabilistic perspective.

Consider a standard uniform distribution ($U(0,1)$), which allows you to randomly pick any real number between 0 and 1 with equal probability. Imagine we keep drawing random numbers from this distribution. The 1st value we pick must be the largest value we have observed as of yet, just like the first competitor to complete a time trial will always be the leader in the competition, even if only temporarily. Therefore, the probability that it is the largest value, $P(largest)$, is 1. Thereafter, in order for the 2nd value we draw to be the largest, it has to be larger than the 1st value; since both values are drawn randomly, it has $P(largest)$ = $\frac{1}{2}$. Likewise, in order for the 3rd value to be the largest, it has to be larger than the 1st and 2nd draws, giving it a  $P(largest)$ of $\frac{1}{3}$.  This logic continues, meaning the probability that the nth draw is larger than all previous draws is $\frac{1}{n}$. Under this model, we expect that we will see a lot of “bests” early on, but fewer and fewer as time goes on. This concept makes intuitive sense when applied our topic. When a sport is new, an athlete is competing with only a few predecessors. Later on, when competing to be the best ever, an athlete may be competing with the same number of contemporaries, but also competing against many more “ghosts” from the pasts who could have finished their careers decades ago. This simple-and perhaps over simplistic-model assumes that the distribution remains the same; in other words, it assumes that there is no growth over time and that athletes do not improve over tine.

Conversely, we can consider the other extreme that holds in a different domain. [\textcolor{blue}{Moore's Law}](https://www.technologyreview.com/2020/02/24/905789/were-not-prepared-for-the-end-of-moores-law/) states “the number of transistors on integrated circuits doubles about every two years.” In that case, as long as Moore’s Law holds, we systematically expect the latest generation of computers to be much faster than all previous generations.^[It is worth noting that by invoking doubling, Moore’s Law describes exponential growth, but even linear growth could suffice to expect new computers to be faster than old computers. Consider a variant of Moore’s Law that would instead say “the number of transistors on integrated circuits increases by 1000 every two years.”]

A linear variant of Moore’s law may describe the progress of professional athletes. Over time, sports has become a bigger business, and training methods have improved. J.J. Reddick caused a bit of a stir when he said, on his podcast *The Old Man & the Three*, that Bob Cousy played against “plumbers and firemen.” As a stylized model, let’s imagine that in the first year, players are drawn from a standard Gaussian distribution ($N(0,1)$). In any subsequent year $i$, the players are drawn from a $N(c(i-1), 1)$ distribution. The distribution shift  means that the expected average ability increases by $c$ units each year. If c is very large, we would expect each year to be much better than every previous year. This model is still over simplistic in that it assumes a constant linear increase in skill, when in fact, skill growth may accelerate or decelerate. Additionally, rule changes, new equipment, or revolutionary strategies can represent a larger one-time spike. It seems plausible that new swimsuits in 2008-2009 (described in more detail in [Revolutions in Sports]) represented a one-time drastic boost, but that advantage was soon removed when the suits were banned, leaving the original moderate improvement trend without the spike from equipment. Furthermore, improving competition may obscure greatness, as an outlier in an earlier era may be merely excellent in a later one. On a podcast called *The Hoopaholics", Brooklyn Nets guard Spencer Dinwiddie said that he would be "God" if he were dropped in the 1960's and played with NBA players from then.  

Let’s examine this simple model. Imagine that each year for 100 years, $n$ new athletes turn professional. Each one of those $n$ new professionals has the same probability of exceeding all values observed in prior years, given by:

$P_{i1} = \int_{-\infty}^{\infty}\prod_{j=1}^{i-1}\prod_{k=1}^{n}F_{x_{jk}}(a)f_{x_{i1}}(a)da$

Where $F_(X_jk)$ represents the CDF of a value in year j, and $f_(X_i1)$ represents the pdf of a value in year $i$. The probability that no value in year $i$ is greater than all values observed in previous years is therefore $(1-P_{i1})^n$ , so the probability the greatest year appears in year $i$ is $1-(1-P_{i1})n$.

Chart 1 shows the probability that an athlete debuting each year will have the highest value observed so far under different values of $c$ and $n$. 

```{r}

library(wesanderson)

#initialize results for 
#N= 10, 100, 1000
#C = 0, .01, .05, .1, .5

results_n_10_c_0 = c(1, rep(0, 99))
results_n_100_c_0 = c(1, rep(0, 99))
results_n_1000_c_0 = c(1, rep(0, 99))


results_n_10_c_01 = c(1, rep(0, 99))
results_n_100_c_01 = c(1, rep(0, 99))
results_n_1000_c_01 = c(1, rep(0, 99))

results_n_10_c_05 = c(1, rep(0, 99))
results_n_100_c_05 = c(1, rep(0, 99))
results_n_1000_c_05 = c(1, rep(0, 99))

results_n_10_c_10 = c(1, rep(0, 99))
results_n_100_c_10 = c(1, rep(0, 99))
results_n_1000_c_10 = c(1, rep(0, 99))

results_n_10_c_50 = c(1, rep(0, 99))
results_n_100_c_50 = c(1, rep(0, 99))
results_n_1000_c_50 = c(1, rep(0, 99))

#This is the main function of interest. We are looking at the product of the CDFs
# of previous generations (with lower mean values) multiplied n times, where n is
# the number of players per generation
# integrated over the pdf of the skill of a player in the current generation

integrand <- function(x, i, n, inc_const) {
  (prod(pnorm(x, mean = seq(from =0, to = inc_const*(i-2), length.out=i-1)))^n)*dnorm(x, mean= inc_const*(i-1))  
  
}

#Run this integration for 100 years/generations, each year incrementing the skill level and number of previous generations
#Note that since sigma=1, we're just integrating from  mean - 5 to mean +5, where the mean is of the current generation
#First we calculate the probability 1 player will be the best, then from there, we calculate the probability that at least 1 player in the 
#current generation will be better than all players of all previous generations
for(i in 2:100){
#print(i)
  
result_10_0 = integrate(Vectorize(integrand), lower = -5, upper = 5, subdivisions = 100000L, i=i, inc_const=0, n=10, rel.tol = .5e-11,  abs.tol = 0)$value
result_100_0 = integrate(Vectorize(integrand), lower = -5, upper = 5, subdivisions = 100000L, i=i, inc_const=0, n=100, rel.tol = .5e-11,  abs.tol = 0)$value
result_1000_0 = integrate(Vectorize(integrand), lower = -5, upper = 5, subdivisions = 100000L, i=i, inc_const=0, n=1000, rel.tol = .5e-11,  abs.tol = 0)$value
  
  
result_10_01 = integrate(Vectorize(integrand), lower = (i*.01)-5, upper = (i*.01)+5, subdivisions = 100000L, i=i, inc_const=.01, n=10, rel.tol = .5e-11, abs.tol = 0)$value
result_100_01 = integrate(Vectorize(integrand), lower = (i*.01)-5, upper = (i*.01)+5, subdivisions = 100000L, i=i, inc_const=.01, n=100, rel.tol = .5e-11,  abs.tol = 0)$value
result_1000_01 = integrate(Vectorize(integrand), lower =(i*.01)-5, upper = (i*.01)+5, subdivisions = 100000L, i=i, inc_const=.01, n=1000, rel.tol = .5e-11,  abs.tol = 0)$value

result_10_05 = integrate(Vectorize(integrand), lower = (i*.05)-5, upper = (i*.05)+5, subdivisions = 100000L, i=i, inc_const=.05, n=10, rel.tol = .5e-11, abs.tol = 0)$value
result_100_05 = integrate(Vectorize(integrand), lower = (i*.05)-5, upper = (i*.05)+5, subdivisions = 100000L, i=i, inc_const=.05, n=100, rel.tol = .5e-11,  abs.tol = 0)$value
result_1000_05 = integrate(Vectorize(integrand), lower = (i*.05)-5, upper = (i*.05)+5, subdivisions = 100000L, i=i, inc_const=.05, n=1000, rel.tol = .5e-11,  abs.tol = 0)$value

result_10_10 = integrate(Vectorize(integrand), lower = (i*.1)-5, upper = (i*.1)+5, subdivisions = 100000L, i=i, inc_const=.1, n=10, rel.tol = .5e-11,  abs.tol = 0)$value
result_100_10 = integrate(Vectorize(integrand), lower = (i*.1)-5, upper = (i*.1)+5, subdivisions = 100000L, i=i, inc_const=.1, n=100, rel.tol = .5e-11,  abs.tol = 0)$value
result_1000_10 = integrate(Vectorize(integrand), lower = (i*.1)-5, upper = (i*.1)+5, subdivisions = 100000L, i=i, inc_const=.1, n=1000, rel.tol = .5e-11,  abs.tol = 0)$value

result_10_50 = integrate(Vectorize(integrand), lower = (i*.5)-5, upper = (i*.5)+5, subdivisions = 100000L, i=i, inc_const=.5, n=10, rel.tol = .5e-11,  abs.tol = 0)$value
result_100_50 = integrate(Vectorize(integrand), lower = (i*.5)-5, upper = (i*.5)+5, subdivisions = 100000L, i=i, inc_const=.5, n=100, rel.tol = .5e-11,  abs.tol = 0)$value
result_1000_50 = integrate(Vectorize(integrand), lower = (i*.5)-5, upper = (i*.5)+5, subdivisions = 100000L, i=i, inc_const=.5, n=1000, rel.tol = .5e-11,  abs.tol = 0)$value


#######
results_n_10_c_0[i]= 1-((1-result_10_0)^10)
results_n_100_c_0[i] = 1-((1-result_100_0)^100)
results_n_1000_c_0[i] = 1-((1-result_1000_0)^1000)

results_n_10_c_01[i]= 1-((1-result_10_01)^10)
results_n_100_c_01[i] = 1-((1-result_100_01)^100)
results_n_1000_c_01[i] = 1-((1-result_1000_01)^1000)

results_n_10_c_05[i]= 1-((1-result_10_05)^10)
results_n_100_c_05[i] = 1-((1-result_100_05)^100)
results_n_1000_c_05[i] = 1-((1-result_1000_05)^1000)

results_n_10_c_10[i]= 1-((1-result_10_10)^10)
results_n_100_c_10[i] = 1-((1-result_100_10)^100)
results_n_1000_c_10[i] = 1-((1-result_1000_10)^1000)

results_n_10_c_50[i]= 1-((1-result_10_50)^10)
results_n_100_c_50[i] = 1-((1-result_100_50)^100)
results_n_1000_c_50[i] = 1-((1-result_1000_50)^1000)

}

#Plot the results above. There isn't much action after the curves stabilize, so we're only displaying over 30 generations, not 100
#Why get stability? At some point, the probability that a player from more than G generations ago is still the highest value effectively goes to 0
# at that point, it's only really a (weighted by recency) competition of the past G generations. Think about it, a N(-5, 1) will almost never exceed a N(0, 1), but a N(-.1, 1) may occasionally

wp = wes_palette("Darjeeling1", 5)
plot(results_n_1000_c_50[1:30], ylim=0:1, type='l', col=wp[1], xlab ="Generation", ylab ="GOAT Probability", main = "GOAT Probability with Different N and C" )
lines(1:30, results_n_1000_c_10[1:30], col=wp[2] )
lines(1:30, results_n_1000_c_05[1:30], col=wp[3] )
lines(1:30, results_n_1000_c_01[1:30], col=wp[5] )
lines(1:30, results_n_100_c_50[1:30], lty=2, col=wp[1])
lines(1:30, results_n_100_c_10[1:30], lty=2, col=wp[2])
lines(1:30, results_n_100_c_05[1:30], col=wp[3] , lty=2)
lines(1:30, results_n_100_c_01[1:30], col=wp[5] , lty=2)
lines(1:30, results_n_10_c_50[1:30], lty=3, col=wp[1])
lines(1:30, results_n_10_c_10[1:30], col=wp[2], lty=3 )
lines(1:30, results_n_10_c_05[1:30], col=wp[3], lty=3 )
lines(1:30, results_n_10_c_01[1:30], col=wp[5], lty=3 )
lines(1:30, results_n_1000_c_0[1:30], type='l', lty=1)
#lines(1:100, results_n_100_c_0, col='blue', lwd=4 )
#lines(1:100, results_n_10_c_0, col='blue' )

legend(x =3.2, y = .77, legend=c("c=.5, N=1000", "c=.5, N=100","c=.5, N=10"), col=wp[1], lty=1:3, lwd=2, horiz=T, box.lty=0, cex=.67, bg="transparent")
legend(x =3.2, y = .7, legend=c("c=.1, N=1000", "c=.1, N=100","c=.1, N=10"), col=wp[2], lty=1:3, lwd=2, horiz=T, box.lty=0, cex=.67, bg="transparent")
legend(x =3.2, y = .63, legend=c("c=.05, N=1000", "c=.05, N=100","c=.05, N=10"), col=wp[3], lty=1:3, lwd=2, horiz=T, box.lty=0, cex=.65, bg ="transparent")
legend(x =3.2, y = .56, legend=c("c=.01, N=1000", "c=.01, N=100","c=.01, N=10"), col=wp[5], lty=1:3, lwd=2, horiz=T, box.lty=0, cex=.65, bg ="transparent")
legend(x=12.5, y=.49, legend="c=0, N=1000", col=1, lty=1, lwd=2, horiz=T, box.lty=0, cex=.67, bg ="transparent")



#Let's turn the previous plot on it's head. Instead of keeping c fixed and looking at probabilities over the generations and different values of N
#We can instead fix the generation and plot the probabilities for different values of C
#Below we plot the "stable probability" for different c values from 0 to 1 at generation=100, when N=1000
results_n_1000_c_i = rep(0,101)
for(i in 0:100){
  # print(i)
  result_1000_i = integrate(Vectorize(integrand), lower = (i)-5, upper = (i)+5, subdivisions = 100000L, i=100, inc_const=i/100, n=1000, rel.tol = .5e-11,  abs.tol = 0)$value
  
  results_n_1000_c_i[i+1] = 1-((1-result_1000_i)^1000)
}  

#rate of improvement vs Goat probability - note that this is probably dependent on a ratio of c/sigma, but since sigma=1, we can use the rate itself here 
plot((0:100)/100, results_n_1000_c_i, xlab = "Average Improvement per Generation",  ylab ="Probability of GOAT in 100th Generation", main = "Long Term GOAT Probability (N=1000)", cex=.8)

```

[\textcolor{red}{Add commentary as needed}]

\newpage

## Revolutions in Sports

### Background

Various changes in rule, equipment, and technique have enabled athletes to gain competitive advantages over their predecessors in a plethora of ways. We now elucidate on some of the most prominent ones in the sports that we have covered. 

### 1. Fosbury Flop 

The Fosbury flop is the most predominant technique used in the track and field event high jump. It was introduced by US Olympian and gold medalist Dick Fosbury, who first used the technique at the 1968 Summer Olympics in Mexico City. The technique was revolutionary in the sense that it consists of jumping backwards off the "wrong" foot. While Fosbury was the first athlete to use this technique in 1968, it was quickly adopted, as 28 of the 40 high jumpers used it in the 1972 Munich Olympics. Furthermore, the technique was quickly deemed to be the most effective and efficient way to perform a high jump. Today, the Fosbury flop is the only technique used by Olympic level high jumpers.

The Fosbury flop is a prime example of how advancements in technique and strategy can give athletes advantages over their predecessors. The advent and introduction of the technique has been linked to the rapid world record progression of the high jump in the 1970's and 80's. More information about it can be found [\textcolor{blue}{here}](https://worldathletics.org/news/feature/dick-fosbury-flop)[@Flop].


### 2. Tech Suits

Tech suits burst onto the professional swim scene shortly thereafter the 2004 Summer Olympics in Athens when Speedo asked NASA to help it design swim suits will less drag. In the subsequent years, Speedo released the LZR Racer, which furthermore led to the release of various other aerodynamic tech suits - or "susper-suits" - from prominent swimwear companies. During this period - known as the "shiny suit era" - swimming world records were broken at an unprecedented pace. More specifically, *74* world records were broken by swimmers wearing the LZR Racer suit. Consequently, World Aquatics, formerly known as the Federation Internationale de Natation (FINA), controversially banned tech suits in 2009, which has led a noticeable drop-off in the pace at which world records have been broken [@Tech]. 

Tech suits are a prime example of how advancements in equipment can give athletes advantages over their predecessors. More information about them and their impact(s) on world records can be found [\textcolor{blue}{here}](https://www.swimmingworldmagazine.com/news/can-world-records-supersede-the-super-suit-era/) [@Tech2].

\newpage

## Final Note

If you have further questions about our data collection process or any other aspect of the project, we encourage you to look at our [\textcolor{blue}{GitHub repository}](https://github.com/murray3625/Sports-GOAT-Project). 


\newpage

## References

